Approximation as Differentiation: Notes

### 2025-05-24 notes/stability.tex

\begin{example}[Stability reveals intensional information]
  To see how stability works, consider the following examples of the
  OR operation on the lifted booleans $\mathbb{B}_\bot$. Two functions
  that are stable are the strict and left-short-circuiting
  ORs\footnote{These clauses are shorthand for the graph of the
    function. They are not to be understood as pattern matching
    clauses in a language like Haskell.}:
  \begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{ff}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{ff}&=&\mathsf{ff} \\
      \mathrm{strictOr}&\bot&\_&=&\bot \\
      \mathrm{strictOr}&\_&\bot&=&\bot \\
    \end{array}
    \qquad
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{shortCircuitOR}&\mathsf{tt}&\_&=&\mathsf{tt} \\
      \mathsf{shortCircuitOR}&\mathsf{ff}&x&=&x \\
      \mathsf{shortCircuitOR}&\bot&\_& =& \bot
    \end{array}
  \end{displaymath}
  The function $\mathrm{strictOr}$ is stable. For example, for the
  input-output pair $(\mathsf{tt},\mathsf{ff}), \mathsf{tt}$, the
  minimal input that gives this output is exactly
  $(\mathsf{tt}, \mathsf{ff})$. If we take the approximation
  $\bot \leq \mathsf{tt}$ of the output, then the corresponding
  minimal input is $(\bot, \bot)$.

  The function $\mathrm{shortCircuitOR}$ is also stable. For the
  input-output pair $(\mathsf{tt},\mathsf{ff}),\mathsf{tt}$, the
  minimal input that gives this input is $(\mathsf{tt},\bot)$,
  indicating that the presence of $\mathsf{ff}$ in the second argument
  was not necessary to produce this output. As with
  $\mathrm{strictOr}$, the minimal input required to produce the
  output $\bot \leq \mathsf{tt}$ is again $(\bot,\bot)$.

@@

  In terms of program slicing, however, the $\mathrm{gustave}$
  function does not present a problem. For any particular run of the
  program, there is an unambiguous minimal input that achieves the
  output, no matter that it was not achieved by a sequential
  processing of the input.
\end{example}

The definition of stability has a nice operational meaning, but seems
a little ad-hoc. We can rephrase stability in terms of Galois
connections / adjunctions between principal downsets / slices of the
partial orders.

@@

\begin{enumerate}
\item Posets where the principal downsets are meet-semilattices. And
  posets where the principal downsets are lattices.
\item Stable functions on these.
\item Full and faithful functor into Fam(LatGal), copying the idea
  from automatic differentiation: we formally delineate points and
  their tangent bundles.
\item
  \begin{enumerate}
  \item Posets that are meet semilattices in their downsets, with
    monotone functions form a Tangent Category
  \item To get ``reverse derivatives'', we further assume stability.
  \end{enumerate}
\end{enumerate}

>>>>

\begin{example}[Stability reveals intensional information]
  To see how stability works, consider the following examples of the
  OR operation on the lifted booleans $\mathbb{B}_\bot$. Two functions
  that are stable are the strict and left-short-circuiting
  ORs\footnote{The clauses in these examples are shorthand for the
    graph of the function. They are not to be understood as pattern
    matching clauses in a language like Haskell, where it is not
    possible to match on $\bot$.}:
  \begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{ff}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{ff}&=&\mathsf{ff} \\
      \mathrm{strictOr}&\bot&\_&=&\bot \\
      \mathrm{strictOr}&\_&\bot&=&\bot \\
    \end{array}
    \qquad
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{shortCircuitOR}&\mathsf{tt}&\_&=&\mathsf{tt} \\
      \mathsf{shortCircuitOR}&\mathsf{ff}&x&=&x \\
      \mathsf{shortCircuitOR}&\bot&\_& =& \bot
    \end{array}
  \end{displaymath}
  The function $\mathrm{strictOr}$ is stable. For example, for the
  input-output pair $(\mathsf{tt},\mathsf{ff}), \mathsf{tt}$, the
  minimal input that gives this output is exactly
  $(\mathsf{tt}, \mathsf{ff})$. If we take the approximation
  $\bot \leq \mathsf{tt}$ of the output, then the corresponding
  minimal input is $(\bot, \bot)$. The function
  $\mathrm{shortCircuitOR}$ is also stable. For the input-output pair
  $(\mathsf{tt},\mathsf{ff}),\mathsf{tt}$, the minimal input that
  gives this input is $(\mathsf{tt},\bot)$, indicating that the
  presence of $\mathsf{ff}$ in the second argument was not necessary
  to produce this output. As with $\mathrm{strictOr}$, the minimal
  input required to produce the output $\bot \leq \mathsf{tt}$ is
  again $(\bot,\bot)$.

@@

  In terms of program slicing, however, the $\mathrm{gustave}$
  function does not present a problem. For any particular run (i.e.,
  input-output pair) of the program, there is an unambiguous minimal
  input that achieves the output, no matter that it was not achieved
  by a sequential processing of the input.
\end{example}

The definition of stability has a nice operational meaning, but seems
a little ad-hoc. We can rephrase stability in terms of Galois
connections / adjunctions between principal downsets / slices of the
partial orders. The alternative presentation also has the advantage
that it explicitly defines the stability witness as ``backwards''
function from approximations of the output to approximations of the
input.

@@

For what follows\footnote{Need a better way of motivating this.}, it
will be useful for the principal downsets to be bounded lattices, so
that we are guaranteed to have meets and joints of approximations of
points. Note that the principal downsets always have a top element
(the original element), but now we are also requiring a bottom
element, as well as finite meets and joins.

FIXME: is this a good name? Maybe just $\mathbf{Stable}$ or something
like that?

FIXME: if $x \leq x'$ then $\downset{x} \subseteq \downset{x'}$ and I
assume that the inclusion preserves lattice structure?

\begin{definition}
  An \emph{approximation set} is a partial order such that every
  principal downset $\downset{x}$ is a bounded lattice.

  A morphism of approxmation sets is monotone function whose
  restriction to each slice has a left adjoint.
\end{definition}

Note that, by standard properties of Galois connections/adjoints, the
pairs $g_x \vdash f_x$ preserve the finite joins and meets respectively.

\begin{proposition}
  Approximation sets and their morphisms form a category.
\end{proposition}

FIXME: what properties does this category have? I think it has finite
products, and probably finite coproducts?

FIXME: there is a full and faithful functor from approximation sets to
$\Fam(\LatGal)$.
